\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-09-23 chapters/wcble.tex
started wcble introduction

\todo{mention published}
\section{Introduction}

Single-trial ERP latency jitter introduces a problem in ERP analysis:
the \textit{smearing effect} occurs when aggregating over multiple identical
signals with different latencies.
A familiar example of this happens when averaging over jittered epochs for ERP
analysis.
In the presence of jitter, the shape and amplitude of the average do not
reflect the properties of the original signal.
As the smearing effect reduces the amplitude of the average,
it negatively impacts the signal-to-noise ratio (SNR).
The smearing effect equally impacts the SNR of information captured by a
classifier's parameters when training with a procedure that does not account for
this jitter, and thus also affects its performance~\cite{Thompson2012}.
Hence, effective latency estimation and jitter compensation can contribute to higher
decoding performance.
%In this work, we propose a latency estimation and alignment strategy based on
%the method proposed by \cite{Thompson2012}.

\todo{General introduction of why we're doing this}

\subsection{ERP latency jitter}

\subsubsection{Neural origins of latency jitter in the time domain}
While the classic ERP model assumes a constant ERP response over trials,
research has shown that cortical evoked potentials show significant
trial-to-trial jitter~\cite{truccolo2002trial}.
Experiments show that latency jitter in ERPs is the result of the interacion of
the evoked activity with ongoing dynamics in the brain~\cite{Hasenstaub2007,
	kisley1999trial, Curto2009, Arieli1996}.

The latency of an evoked potential is a combination of the modulation of oscillating activity that is
time-locked, but not necessarily phase-locked to stimulus onset, and a
delay caused by ongoing neurophysiological processes that are not necessarily related to the task.
\todo{Check this, \cite{Stokes2016,Mouraux2008}}
\todo{Read references and check if neural origins and
	generation model justify approach}

\subsubsection{Implications for ERP analysis}
The presence of latency jitter introduces multiple issues to ERP analysis, the
most prominent of which is the smearing effect as illustrated
in~\autoref{fig:smearing}. In ERP analysis, single-trial ERPs are usually
averaged per condition to cope with the unfavorable signal-to-noise ratio  When
averaging over multiple identical signals with jittered latencies, the shape and amplitude of the average do not reflect the properties
of the original signal. Because the amplitude of the average is lower, the
signal-to-noise ratio will also be negatively impacted.
More generally, the smearing effect leads to difficulties interpreting ERP
results and making inferences about amplitude or latency effects, because these
are entangled when using averaging~\cite{woody1967characterization, Pfefferbaum1980, McDowell2003, Verleger2005,
	Roth2007, Walhovd2008, Poli2010, Luck2014}\todo{Review relevance of
	these references}.

\todo{illustrate smearing effect}
%\begin{figure}
%  \todo{smearing effect}
%	\caption{The smearing effect: averaging over jittered waveforms leads to a spread out
%		average with a lower amplitude.}
%	\label{fig:smearing}
%\end{figure}

\subsubsection{Benefits of latency estimation}
A common approach to counteract the smearing effect is correcting single-trial
ERPs for jitter before averaging.\todo{reference} This can be done by estimating the latency of each
single-trial ERP, and aligning them before averaging.
Latency estimation has multiple advantages:
\begin{enumerate*}[label={(\arabic*)}]
	\item Latency-based ERP alignment can help reveal the true shape and
	properties like peak amplitude and latency of ERPs across
	experimental conditions. These are variables of interest for
	testing electrophysiological an behavioral hypotheses.
	Additionally, latency estimation allows for the extraction of
	other descriptors from a set of single-trial ERPs, like
	amplidute or latency variance~\cite{Hultsch2004}, which can
	also be of interest as studied variables. For instance, in
	\cite{Saville2014}, the inter-trial latency variability itself
	is determined as a correlate of working memory performance.
	\item Counteracting the smearing effect will boost the signal-to-noise
	ratio of averaged ERPs by increasing the amplitude of the
	ERP signal. This in turn can reveal smaller or more jittered ERP components
	which are otherwise obscured or blurred due to desynchronization.
	These deblurred templates can then be used for analysis of the ERP
	waveform, or as templates of the signal of interest in
	BCIs~\cite{arico2014influence}.
	%this enables decoders to take into account the relatively low amplitude
	%earlier components, leading to applications in settings where
	%information from other components than the P300 need to be exploited,
	%like covert visual attention paradigms~\cite{arico2014influence,
	%hardiansyah2020single}.\todo[inline]{rework this to latency features
	%because smaller visual components might not really be revealed
	%since they're already time-locked}
	\item Finally, the introduction of latency estimation and alignment generaly
	improves BCI classification performance. Single-trial latencies can be
	used as features to improve classification~\cite{hardiansyah2020single}.
	Furthermore, compensating for inter-trial differences, especially when
	extended to inter-subject differences, can aid in transfer learning or
	generalization across subjects and protocols\cite{Iturrate2014}.
\end{enumerate*}
\todo{Read, summarize and cite BCI	references from bibliography}

@@

\subsection{Single-component approaches}
The most straight-forward method to estimate single-trial latency
peak-picking. Here, the latency of an ERP is determined as the
time-point of it's maximum (or minimum) amplitude with relaion to the stimulus
onset. While straight-forward, this method
does not perform well in low or negative SNR conditions, unless combined with
filtering. Filtering can suppress the noise contaminating EEG trials, lowering
the risk of picking a noise peak instead of a true peak of the ERP. Several filtering approaches have been developed to be used in
conjuction with peak-picking, in the time domain~\cite{Carlton1980, McCarthy1981,
	Karjalainen1999, Nishida1999,  Sparacino2002, Georgiadis2005,
	D’Avanzo2011} and in the time-frequency
domain~\cite{Quiroga2003, wang2007single,Hu2010,arico2014influence}. Filtering can
also be done in the spatial domain. The spatial filter can be constructed
using PCA~\cite{Ouyang2017}, ICA~\cite{townsend1999analyzing, jung2001analysis,
	Milne2011}, or as a spatial LCMV beamformer~\cite{TREDER2016279}.

A generally more preferred approach than
peak-picking is template matching. Woody's algorithm~\cite{woody1967characterization} is a simple and performant latency
estimation method that is still being used and is considered amongst the more
performant techniques~\cite{Ouyang2017}. It starts with the average ERP as a
rough estimation of the aligned template ERP, and iteratively refines this
template by determining at each step the time point of maximum
cross-correlation of this template with all single-trial ERPs and aligning them
on these time points to form the next template.
Similarily, cross-correlation template matching can be combined with filtering.
\cite{souloumiac2013improved} uses XDAWN to construct spatial filter before
template matching the spatially filtered signal. \cite{Iturrate2014} uses a
spatiotemporal filter. Instead of aligning the
latencies to obtain the template for the next iteration, weighted averaging can
also be used~\cite{gasser1983selavco}.

Despite good performance, the iterative scheme of Woody's algorithm has a risk of converging to a local
optimum. To avoid this, a fitness function can be defined to jointly optimize
the set of single-trial ERP latencies, which can be optimized with a genetic algorithm~\cite{Pelo2018}.

@@

\subsection{Multi-component approaches}
The abovementioned methods do not take into account the fact that  ERPs are multi-component signals. Multiple components can be time-locked to different events or neurophysiological proccess and hence are not necessarily
time-locked to each other. They each can have different latency
variabilities. The presence of more than one component can hamper the
performance of some algorithms, and it often only possible to extract only the
latencies of the largest component present~\cite{Ouyang2017}. When aligning trials
to one component and averaging, this will introduce blurring in the other
components~\cite{Ouyang2020}.

Most of the algorithms above can be adapted to work on multiple ERP components
by carefully selecting peaks or pre-determined regions of interest.
\cite{hardiansyah2020single} uses an SNR boosting method, applied to a specific
time-window for each ERP component to extract a latency per component.
Methods based on spatial decomposition filters like PCA, ICA or XDAWN can be leveraged
to utilize multiple of their components as filters, resulting in multiple time
series where peack picking or Woody's algorithm can be applied. In general,
however, these algorithms lack an integrated approach to deal with multiple
ERP components. A clear evaluation is lacking of the performance of these
methods in
multi-component settings is lacking.

Some algorithms separate components into stimulus and response locked component clusters based on their
latency distributions~\cite{jung2001analysis, Takeda2008,
	Zhang1998, Yin2009}. These algorithms do, however, require a response
button trigger signal as input to determine the response time, which is
not always present or applicable in an ERP experiment, especially in
the case of BCIs.
The RIDE algorithm~\cite{Ouyang2011, Ouyang2015,
	Wang2015, Ouyang2016, Ouyang2020} is able to do this without entirely depending on
the response time, and can separate stimulus-locked, response-locked,
and multiple central component clusters. Others, like spatial filtering methods that use
decomposition methods to determine the spatial filter can be be modified to separate
components.~\cite{Ouyang2017}.

@@

Finally, the smearing problem can also be overcome by algorithms that are not
based on ERP component latency estimation, but which use time warping-based methods
to enhance the shape of the averaged template ERP, like Dynamic Time
Warping~\cite{486255, wang2001warp, Zoumpoulaki2015}, Correlation-Optimized
Warping~\cite{Skov2006}, or Fast Variational
Alignment~\cite{flotho2021fast}. Since these methods work
fundamentally different from the latency estimation methods, and cannot
directly be used to extract latency features, we will not further consider them
in this work.

@@

The most straightforward latency estimation method is \textit{peak picking}.
Peak picking determines the latency of an ERP as the time point of its maximum
(or minimum for negative components) amplitude relative to stimulus
onset.
While easy to implement, this method does not perform well in low or negative
SNR conditions because of the risk of picking a noise-related peak, unless
combined with strong filtering in the space, time or frequency
domains~\cite{Avanzo2011,Arico2014,TREDER2016279,Ouyang2017}.
The aforementioned analysis and performance prediction method proposed
by \cite{Arico2014} estimates the single-trial P3 latencies of attended
epochs by P3 peak-picking after filtering in the time-frequency domain, and
corrects for jitter by temporally aligning all target epochs to the average
latency, i.e., by shifting the epochs in time such that the P3 peaks all fall
at the same time instance.
Due to the risk of picking noise peaks, template matching is generally
preferred.
Woody's algorithm~\cite{woody1967characterization} is a simple and performant
latency estimation method that is considered to be among
the most performant ones~\cite{Ouyang2017}.
Woody's scheme starts with a template that approximates the expected response,
e.g. the average target ERP, and iteratively enhances the SNR of this template by determining at
each iteration the time point of maximum cross-correlation of the template
with all single-trial ERPs, and aligning them based on these latencies to form
the template for the next iteration.
The final, enhanced template can then be cross-correlated again to retrieve an
accurate estimation of single-trial latencies.
Similar to peak-picking, cross-correlation template matching can be combined
with a spatial filter such as XDAWN~\cite{souloumiac2013improved} or a
spatiotemporal filter \cite{Iturrate2014}.
More recent work achieves good performance in latency estimation using residue
iteration~\cite{Ouyang2017}, genetic
algorithms~\cite{Pelo2018}, a hidden process model~\cite{Kim2020} or
graph-based network analysis~\cite{Dimitriadis2018}.

When multiple clusters of ERP components are not time-locked to
each other, aligning to the latency of one cluster might introduce more jitter
in other component clusters, further decreasing the SNR~\cite{Ouyang2020}.

Finally, algorithms that use time
warping-based methods instead of explicit ERP component latency estimation can
also overcome the smearing problem.
Examples that have been applied to ERPs are Dynamic Time
Warping~\cite{Casarotto2005,486255,wang2001warp,Zoumpoulaki2015},
Correlation-Optimized Warping~\cite{Skov2006}, or Fast Variational
Alignment~\cite{Flotho2021}. Since these methods are fundamentally
different from latency estimation methods,
we will not further consider them in this work.

@@

Most of these methods suffer from a common drawback: they cannot be used in a
decoding scheme to improve performance for incoming epochs with unknown
class labels that contain either a target or non-target ERP response.
They can be used offline on a set of labeled epochs for testing hypotheses
concerning latency and jitter, for aligning templates, or for BCI performance
prediction, but not for ERP classification.
While some of the aforementioned methods could be adapted to perform
classification tasks, few studies investigate how to exploit this latency estimation
for jitter-resistant decoding.
\cite{hardiansyah2020single} incorporated single-trial latencies in
classification by peak-picking within a given
ERP time window, unaware of the class of the epoch under investigation.
The Classifier-Based Latency Estimation (CBLE) algorithm introduced
by \cite{Thompson2012} also explicitly applies latency estimation in a
decoding setting.
\cite{Thompson2012} initially formulated CBLE as an off-line performance prediction method.
Later, its output was successfully adapted to compensate for jitter to improve decoder performance%
~\cite{Mowla2017,Zisk2022}.

Time series classification algorithms~\cite{Abanda2019}
that are robust to jitter can be used in a decoding setting,
but, in general, have scarcely been applied to ERP decoding.
Data augmentation involving jittering the training
data~\cite{Krell2018,Zisk2022} and Riemannian Geometry methods using spatial
covariances as features~\cite{Aydarkhanov2020} have both been shown
to perform well in the presence of ERP jitter.
In this work, we opted to apply CBLE because it has successfully been applied to
classify jittered ERPs.
We adapt the CBLE to an iterative method akin to Woody's scheme that can better
compensate for jitter, to improve covert VSA decoding performance.

\todo{remove lcmv}
In this work, we introduce a new latency method that leverages data-driven,
joint spatiotemporal filtering based on the LCMV beamformer~\cite{Vliet2016}. The algorithm
integrates the spatiotemporal filtering in Woody's template matching iteration
scheme and can easily be extended to multi-component settings.
Our method attempts to model the spatial and temporal activation of
each ERP component, as well as it's latency distribution, and takes into acount the
spatial and temporal correlation of the EEG background noise.

@@

CBLE, summarized in Algorithm~\ref{alg:cble}, works by training a
\textit{first-stage} classifier $\mathcal{C}(\theta,f)$
defined within a time period $[s_1:s_2]$ with a set of parameters $\theta$ and a
decision function $f(\mat{X}[:,s_1:s_2],\theta) \rightarrow
y$ outputting a classification score $y\in\mathbb{R}$ for a given epoch $\mat{X}$,
such that
\begin{equation}
  \theta = \mathrm{train}_\mathcal{C}(\{\mat{X}_n^\mathrm{train}[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
  \label{eq:cble-train}
\end{equation}
Then,
%instead of applying the classifier directly to an unseen, cropped
%testing epoch $\mat{X}[:,s_1:s_2]$ and obtaining a single score value $y$,
$f$ can be applied to all (possibly overlapping) slices of length $s_2-s_1$ of
an epoch $\mat{X}$, resulting in a vector of score values
$\mathbf{y}=[y_1\,\ldots\,y_R]^T \in\mathbb{R}^R$ such that
\begin{equation}
  y_s = f(\mat{X}[:,s:s+(s_2-s_1)],\theta)\quad \forall s\in 1,\ldots,R
	\label{eq:score}
\end{equation}
with $R = S-(s_2-s_1)$.
To leverage CBLE for ERP classification, the score vectors $\mathbf{y}$ can be
arranged in matrices $\mat{Y}^\mathrm{train}\in\mathbb{R}^{N\times R}$ and $\mat{Y}^\mathrm{test}\in\mathbb{R}^{M\times R}$.
These can be further classified by training a \textit{second-stage} classifier on
$\mat{Y}^\mathrm{train}$ and class labels $\mathbf{l}^\mathrm{train}$.
However, the resulting score-over-time vectors per epoch still suffer from jitter.
For classification, we follow the approach of \cite{Mowla2017}, using a
maximum-level hierarchical Daubechies-4 wavelet transform to reduce
dimensionality before classification with the second-stage classifier.
In the CBLE-decoder, it is this wavelet transform that decreases the sensitivity
to latency differences, actively compensating for ERP latency jitter.

When using a simple spatiotemporal linear classifier as first-stage classifier,
CBLE is equivalent to the first iteration of Woody's algorithm with the
spatiotemporal classifier weights as template.
\cite{Thompson2012}, \cite{Mowla2017} and \cite{Mowla2020} show
that CBLE is relatively independent of the first-stage classifier for BCI
accuracy prediction and for ERP classification.
Therefore, we opt to use the variant of Linear Discriminant Analysis with block-Toeplitz
regularized covariance matrix (tLDA) proposed by \cite{Sosulski2022}, as the first-stage
classifier and logistic regression as second stage.

\subsection{Robust CBLE latency features}
\label{sec:robust-latency}
The ERP decoding method based on CBLE introduced by \cite{Mowla2017} does
not make use of the extracted latencies, only passing score matrix $\mat{Y}$ on
to the second-stage classifier.
%Furthermore, previous CBLE works~\cite{Thompson2012,Mowla2020} only used these
%latencies to correlate them with neurophysiological processes or to predict
%decoder performance.
We noticed that, while CBLE performance was unaffected, the classification
performance of our proposed method can be improved if the estimated
latencies are also made available as features to the second-stage classifier,
after a square transform for linear separability~\cite{Thompson2012}.

>>>>

This chapter, except Sections~\ref{sec:wcble/intro},
\ref{sec:wcble/literature/single-comp} and
\ref{sec:wcble/literature/multi-comp}
\todo{properly assess which paragraphs were published and which not}

\section{Introduction}
\label{sec:wcble/intro}

Visual oddball \acp{bci} establish communication with paralyzed
individuals by decoding \acp{erp} obtained from time-modulated flashing
stimuli.
These \acp{erp} consist of multiple \ac{erp}-components, peaks and troughs in
\ac{erp} waveform.
These components each have their own distinct amplitude and latency.
\Ac{erp} component latency can reflect a wide range of cognitive,
neurophysiological and clinical properties, that can be confounding factors in
the development of \acp{bci}.\todo{cite}
Latency is often assessed as the latency of \ac{erp} components in a subject
average or grand average.
Latencies of non-averaged, single-trial \ac{erp} waveforms could capture even more
relevant information, yet they are very hard to measure due to low \ac{snr} of
single trial \acp{erp}.
Within session variability of \ac{erp} latency, or \ac{erp} latency
\emph{jitter} has been shown to influence \ac{bci}
accuracy~\cite{Thompson2012}.
This effect is especially prominent when bringing \ac{bci} development from the
lab to the clinical setting with patients in the intended target population.
\textcite{Zisk2021} showed that patients with \ac{als} have significantly
increased ERP jitter.
\textcite{Arico2014} show that latency jitter negatively affects \ac{bci}
performance, and disproportionally so when the user is not directly gazing at
intended targets, such as is the case for for gaze-impaired patients.

\subsection{\Ac{erp} latency jitter}

\subsubsection{Neural origins of latency jitter}
While the classic \ac{erp} model assumes a constant \ac{erp} response over trials,
research has shown that cortical evoked potentials show significant
trial-to-trial jitter~\cite{Truccolo2002}.
Experiments show that latency jitter in \ac{erp}s is the result of the interacion of
the evoked activity with ongoing dynamics in the brain~\cite{Hasenstaub2007,
	Kisley1999, Curto2009, Arieli1996}.
The latency of an evoked potential is a combination of the modulation of oscillating activity that is
time-locked, but not necessarily phase-locked to stimulus onset, and a
delay caused by ongoing neurophysiological processes that are not necessarily related to the task.
\todo{Check this, \cite{Stokes2016,Mouraux2008}}
\todo{Read references and check if neural origins and
	generation model justify approach}

\subsubsection{Implications for \ac{erp} analysis}

The presence of latency jitter introduces multiple issues to \ac{erp} analysis, the
most prominent of which is the smearing effect as illustrated
in~\autoref{fig:smearing}. In \ac{erp} analysis, single-trial \ac{erp}s are usually
averaged per condition to cope with the unfavorable \ac{snr}?
When averaging over multiple identical signals with jittered latencies, the shape
and amplitude of the average do not reflect the properties
of the original signal.
Because the amplitude of the average is lower, the \ac{snr} will also be negatively impacted.
More generally, the smearing effect leads to difficulties interpreting \ac{erp}
results and making inferences about amplitude or latency effects, because these
are entangled when using averaging~\cite{Woody1967, Pfefferbaum1980, McDowell2003, Verleger2005,
	Roth2007, Walhovd2008, Poli2010, Luck2014}\todo{Review relevance of
	these references}.

\todo{illustrate smearing effect}

The smearing effect equally impacts the \ac{snr} of information captured by a
classifier's parameters when training with a procedure that does not account for
this jitter, and thus also affects its performance~\cite{Thompson2012}.
As an illustrative example, think of an \ac{lda} \ac{erp} classifier that uses
class averages over \ac{erp} epochs to construct its between-class scatter
matrix.
These per-class averages will be affected by the smearing effect and will be
less effective templates for \ac{erp} retrieval.
Similarly, the noise model in \ac{lda}'s  assumes stationarity and subtracts
the class-wise averages from the signal to calculate
the within-class scatter matrix, which will then also be affected.
Hence, effective latency estimation and jitter compensation can contribute to higher
decoding performance.

\subsubsection{Benefits of latency estimation}
A common approach to counteract the smearing effect is to correct single-trial
\ac{erp}s for jitter before averaging.\todo{reference}
This can be done by estimating the latency of each
single-trial \ac{erp}, and aligning them before averaging.
Latency estimation has multiple advantages:
\begin{enumerate*}[label={(\arabic*)}]
	\item Latency-based \ac{erp} alignment can help reveal the true shape and
	properties like peak amplitude and latency of \ac{erp}s across
	experimental conditions. These are variables of interest for
	testing electrophysiological an behavioral hypotheses.
	Additionally, latency estimation allows for the extraction of
	other descriptors from a set of single-trial \ac{erp}s, like
	amplidute or latency variance~\cite{Hultsch2004}, which can
	also be of interest as studied variables.
  For instance, in the work of \textcite{Saville2014}, the inter-trial latency
  variability itself is determined as a correlate of working memory performance.
	\item Counteracting the smearing effect will boost the signal-to-noise
	ratio of averaged \ac{erp}s by increasing the amplitude of the
	\ac{erp} signal. This in turn can reveal smaller or more jittered \ac{erp}
  components which would otherwise obscured or blurred due to desynchronization.
	These deblurred templates can then be used for analysis of the \ac{erp}
	waveform, or as templates of the signal of interest in
	BCIs~\cite{arico2014influence}.
	%this enables decoders to take into account the relatively low amplitude
	%earlier components, leading to applications in settings where
	%information from other components than the P300 need to be exploited,
	%like covert visual attention paradigms~\cite{arico2014influence,
	%hardiansyah2020single}.\todo[inline]{rework this to latency features
	%because smaller visual components might not really be revealed
	%since they're already time-locked}
	\item Finally, the introduction of latency estimation and alignment generaly
	improves BCI classification performance. Single-trial latencies can be
	used as features to improve classification~\cite{hardiansyah2020single}.
	Furthermore, compensating for inter-trial differences, especially when
	extended to inter-subject differences, can aid in transfer learning or
	generalization across subjects and protocols\cite{Iturrate2014}.
\end{enumerate*}
\todo{Read, summarize and cite BCI	references from bibliography}

@@

\subsection{Single-component approaches}
\label{sec:wcble/literature/single-comp}

The most straight-forward method to estimate single-trial latency
peak-picking. Here, the latency of an \ac{erp} is determined as the
time-point of it's maximum (or minimum) amplitude with relaion to the stimulus
onset. While straight-forward, this method
does not perform well in low or negative SNR conditions, unless combined with
filtering. Filtering can suppress the noise contaminating EEG trials, lowering
the risk of picking a noise peak instead of a true peak of the \ac{erp}. Several filtering approaches have been developed to be used in
conjuction with peak-picking, in the time domain~\cite{Carlton1980, McCarthy1981,
	Karjalainen1999, Nishida1999,  Sparacino2002, Georgiadis2005,
	D’Avanzo2011} and in the time-frequency
domain~\cite{Quiroga2003, wang2007single,Hu2010,arico2014influence}. Filtering can
also be done in the spatial domain. The spatial filter can be constructed
using PCA~\cite{Ouyang2017}, ICA~\cite{townsend1999analyzing, jung2001analysis,
	Milne2011}, or as a spatial LCMV beamformer~\cite{TREDER2016279}.

A generally more preferred approach than
peak-picking is template matching. Woody's algorithm~\cite{woody1967characterization} is a simple and performant latency
estimation method that is still being used and is considered amongst the more
performant techniques~\cite{Ouyang2017}. It starts with the average \ac{erp} as a
rough estimation of the aligned template \ac{erp}, and iteratively refines this
template by determining at each step the time point of maximum
cross-correlation of this template with all single-trial \ac{erp}s and aligning them
on these time points to form the next template.
Similarily, cross-correlation template matching can be combined with filtering.
\cite{souloumiac2013improved} uses XDAWN to construct spatial filter before
template matching the spatially filtered signal. \cite{Iturrate2014} uses a
spatiotemporal filter. Instead of aligning the
latencies to obtain the template for the next iteration, weighted averaging can
also be used~\cite{gasser1983selavco}.

Despite good performance, the iterative scheme of Woody's algorithm has a risk of converging to a local
optimum. To avoid this, a fitness function can be defined to jointly optimize
the set of single-trial \ac{erp} latencies, which can be optimized with a genetic algorithm~\cite{Pelo2018}.

@@

\subsection{Multi-component approaches}
\label{sec:wcble/literature/multi-comp}
The abovementioned methods do not take into account the fact that  \ac{erp}s are multi-component signals. Multiple components can be time-locked to different events or neurophysiological proccess and hence are not necessarily
time-locked to each other. They each can have different latency
variabilities. The presence of more than one component can hamper the
performance of some algorithms, and it often only possible to extract only the
latencies of the largest component present~\cite{Ouyang2017}. When aligning trials
to one component and averaging, this will introduce blurring in the other
components~\cite{Ouyang2020}.

Most of the algorithms above can be adapted to work on multiple \ac{erp} components
by carefully selecting peaks or pre-determined regions of interest.
\cite{hardiansyah2020single} uses an SNR boosting method, applied to a specific
time-window for each \ac{erp} component to extract a latency per component.
Methods based on spatial decomposition filters like PCA, ICA or XDAWN can be leveraged
to utilize multiple of their components as filters, resulting in multiple time
series where peack picking or Woody's algorithm can be applied. In general,
however, these algorithms lack an integrated approach to deal with multiple
\ac{erp} components. A clear evaluation is lacking of the performance of these
methods in
multi-component settings is lacking.

Some algorithms separate components into stimulus and response locked component clusters based on their
latency distributions~\cite{jung2001analysis, Takeda2008,
	Zhang1998, Yin2009}. These algorithms do, however, require a response
button trigger signal as input to determine the response time, which is
not always present or applicable in an \ac{erp} experiment, especially in
the case of BCIs.
The RIDE algorithm~\cite{Ouyang2011, Ouyang2015,
	Wang2015, Ouyang2016, Ouyang2020} is able to do this without entirely depending on
the response time, and can separate stimulus-locked, response-locked,
and multiple central component clusters. Others, like spatial filtering methods that use
decomposition methods to determine the spatial filter can be be modified to separate
components.~\cite{Ouyang2017}.

@@

Finally, the smearing problem can also be overcome by algorithms that are not
based on \ac{erp} component latency estimation, but which use time warping-based methods
to enhance the shape of the averaged template \ac{erp}, like Dynamic Time
Warping~\cite{486255, wang2001warp, Zoumpoulaki2015}, Correlation-Optimized
Warping~\cite{Skov2006}, or Fast Variational
Alignment~\cite{flotho2021fast}. Since these methods work
fundamentally different from the latency estimation methods, and cannot
directly be used to extract latency features, we will not further consider them
in this work.

@@

The most straightforward latency estimation method is \textit{peak picking}.
Peak picking determines the latency of an \ac{erp} as the time point of its maximum
(or minimum for negative components) amplitude relative to stimulus
onset.
While easy to implement, this method does not perform well in low or negative
SNR conditions because of the risk of picking a noise-related peak, unless
combined with strong filtering in the space, time or frequency
domains~\cite{Avanzo2011,Arico2014,TREDER2016279,Ouyang2017}.
The aforementioned analysis and performance prediction method proposed
by \cite{Arico2014} estimates the single-trial P3 latencies of attended
epochs by P3 peak-picking after filtering in the time-frequency domain, and
corrects for jitter by temporally aligning all target epochs to the average
latency, i.e., by shifting the epochs in time such that the P3 peaks all fall
at the same time instance.
Due to the risk of picking noise peaks, template matching is generally
preferred.
Woody's algorithm~\cite{woody1967characterization} is a simple and performant
latency estimation method that is considered to be among
the most performant ones~\cite{Ouyang2017}.
Woody's scheme starts with a template that approximates the expected response,
e.g. the average target \ac{erp}, and iteratively enhances the SNR of this template by determining at
each iteration the time point of maximum cross-correlation of the template
with all single-trial \ac{erp}s, and aligning them based on these latencies to form
the template for the next iteration.
The final, enhanced template can then be cross-correlated again to retrieve an
accurate estimation of single-trial latencies.
Similar to peak-picking, cross-correlation template matching can be combined
with a spatial filter such as XDAWN~\cite{souloumiac2013improved} or a
spatiotemporal filter \cite{Iturrate2014}.
More recent work achieves good performance in latency estimation using residue
iteration~\cite{Ouyang2017}, genetic
algorithms~\cite{Pelo2018}, a hidden process model~\cite{Kim2020} or
graph-based network analysis~\cite{Dimitriadis2018}.

When multiple clusters of \ac{erp} components are not time-locked to
each other, aligning to the latency of one cluster might introduce more jitter
in other component clusters, further decreasing the SNR~\cite{Ouyang2020}.

Finally, algorithms that use time
warping-based methods instead of explicit \ac{erp} component latency estimation can
also overcome the smearing problem.
Examples that have been applied to \ac{erp}s are Dynamic Time
Warping~\cite{Casarotto2005,486255,wang2001warp,Zoumpoulaki2015},
Correlation-Optimized Warping~\cite{Skov2006}, or Fast Variational
Alignment~\cite{Flotho2021}. Since these methods are fundamentally
different from latency estimation methods,
we will not further consider them in this work.

@@

Most of these methods suffer from a common drawback: they cannot be used in a
decoding scheme to improve performance for incoming epochs with unknown
class labels that contain either a target or non-target \ac{erp} response.
They can be used offline on a set of labeled epochs for testing hypotheses
concerning latency and jitter, for aligning templates, or for BCI performance
prediction, but not for \ac{erp} classification.
While some of the aforementioned methods could be adapted to perform
classification tasks, few studies investigate how to exploit this latency estimation
for jitter-resistant decoding.
\cite{Hardiansyah2020} incorporated single-trial latencies in
classification by peak-picking within a given
\ac{erp} time window, unaware of the class of the epoch under investigation.
The Classifier-Based Latency Estimation (CBLE) algorithm introduced
by \cite{Thompson2012} also explicitly applies latency estimation in a
decoding setting.
\cite{Thompson2012} initially formulated CBLE as an off-line performance prediction method.
Later, its output was successfully adapted to compensate for jitter to improve decoder performance%
~\cite{Mowla2017,Zisk2022}.

Time series classification algorithms~\cite{Abanda2019}
that are robust to jitter can be used in a decoding setting,
but, in general, have scarcely been applied to \ac{erp} decoding.
Data augmentation involving jittering the training
data~\cite{Krell2018,Zisk2022} and Riemannian Geometry methods using spatial
covariances as features~\cite{Aydarkhanov2020} have both been shown
to perform well in the presence of \ac{erp} jitter.
In this work, we opted to apply CBLE because it has successfully been applied to
classify jittered \ac{erp}s.
We adapt the CBLE to an iterative method akin to Woody's scheme that can better
compensate for jitter, to improve covert VSA decoding performance.

\todo{remove lcmv}
In this work, we introduce a new latency method that leverages data-driven,
joint spatiotemporal filtering based on the LCMV beamformer~\cite{Vliet2016}. The algorithm
integrates the spatiotemporal filtering in Woody's template matching iteration
scheme and can easily be extended to multi-component settings.
Our method attempts to model the spatial and temporal activation of
each \ac{erp} component, as well as it's latency distribution, and takes into acount the
spatial and temporal correlation of the EEG background noise.

@@

CBLE, summarized in Algorithm~\ref{alg:cble}, works by training a
\textit{first-stage} classifier $\mathcal{C}(\theta,f)$
defined within a time period $[s_1:s_2]$ with a set of parameters $\theta$ and a
decision function $f(\mat{X}[:,s_1:s_2],\theta) \rightarrow
y$ outputting a classification score $y\in\mathbb{R}$ for a given epoch $\mat{X}$,
such that
\begin{equation}
  \theta = \mathrm{train}_\mathcal{C}(\{\mat{X}_n^\mathrm{train}[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
  \label{eq:cble-train}
\end{equation}
Then,
%instead of applying the classifier directly to an unseen, cropped
%testing epoch $\mat{X}[:,s_1:s_2]$ and obtaining a single score value $y$,
$f$ can be applied to all (possibly overlapping) slices of length $s_2-s_1$ of
an epoch $\mat{X}$, resulting in a vector of score values
$\mathbf{y}=[y_1\,\ldots\,y_R]^T \in\mathbb{R}^R$ such that
\begin{equation}
  y_s = f(\mat{X}[:,s:s+(s_2-s_1)],\theta)\quad \forall s\in 1,\ldots,R
	\label{eq:score}
\end{equation}
with $R = S-(s_2-s_1)$.
To leverage CBLE for \ac{erp} classification, the score vectors $\mathbf{y}$ can be
arranged in matrices $\mat{Y}^\mathrm{train}\in\mathbb{R}^{N\times R}$ and $\mat{Y}^\mathrm{test}\in\mathbb{R}^{M\times R}$.
These can be further classified by training a \textit{second-stage} classifier on
$\mat{Y}^\mathrm{train}$ and class labels $\mathbf{l}^\mathrm{train}$.
However, the resulting score-over-time vectors per epoch still suffer from jitter.
For classification, we follow the approach of \cite{Mowla2017}, using a
maximum-level hierarchical Daubechies-4 wavelet transform to reduce
dimensionality before classification with the second-stage classifier.
In the CBLE-decoder, it is this wavelet transform that decreases the sensitivity
to latency differences, actively compensating for \ac{erp} latency jitter.

When using a simple spatiotemporal linear classifier as first-stage classifier,
CBLE is equivalent to the first iteration of Woody's algorithm with the
spatiotemporal classifier weights as template.
\cite{Thompson2012}, \cite{Mowla2017} and \cite{Mowla2020} show
that CBLE is relatively independent of the first-stage classifier for BCI
accuracy prediction and for \ac{erp} classification.
Therefore, we opt to use the variant of Linear Discriminant Analysis with block-Toeplitz
regularized covariance matrix (tLDA) proposed by \cite{Sosulski2022}, as the first-stage
classifier and logistic regression as second stage.

\subsection{Robust CBLE latency features}
\label{sec:robust-latency}
The \ac{erp} decoding method based on CBLE introduced by \cite{Mowla2017} does
not make use of the extracted latencies, only passing score matrix $\mat{Y}$ on
to the second-stage classifier.
%Furthermore, previous CBLE works~\cite{Thompson2012,Mowla2020} only used these
%latencies to correlate them with neurophysiological processes or to predict
%decoder performance.
We noticed that, while CBLE performance was unaffected, the classification
performance of our proposed method can be improved if the estimated
latencies are also made available as features to the second-stage classifier,
after a square transform for linear separability~\cite{Thompson2012}.

These latency features give rise to the following issue when
classifying unseen data.
The CBLE latency estimate is defined only for target epochs as
$s_\mathrm{target}=\argmax_s{y_s}$.
This is the point in time where the target class reaches largest separation
from the background noise and non-target class, indicating the target \ac{erp} is
most likely to occur here.
However, in a classifier test phase, it is not known a priori whether an
unseen epoch is a target or a non-target epoch.
This problem is solved by defining an estimated latency per class
$s_\mathrm{target}$ and $s_\mathrm{non-target}$ for every epoch,
regardless of its actual class.
The estimated class latencies can then be used as features  for training and
testing the second-stage classifier.
This way, the latencies of the testing data can be presented to the second-stage
classifier without knowledge of the testing data class labels, making them
useful in decoding.
